pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-29 article.tex
Updates from ShareLaTeX

Phylogenies are commonly used to explore evolutionary hypotheses.
The mode of diversification is one such hypothesis.
Phylogenies constructed from DNA alignments are commonly fit to
diversification models, or the other way around. For example,
[Etienne & Rosindell 2012] fit four bird phylogenies to the protracted
diversification model.

There are plenty of diversification models: constant-birth [Yule, 19..],
birth-death [Nee et al., 19..], time-dependent [?], diversity 
dependent [Etienne & ?, approx 2012], protracted 
birth-death [Rosindell & Etienne, 2011][Etienne & Rosindell, 2012],
multiple-birth death [Laudanno, Bilderbeek & Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno & Etienne, in preparation]
punctuated equilibrium aka early burst [?Harmon],
dependent on a binary trait [BISSE model, ?Maddison], 
continuous trait [HISSE model],
concealed state [?MUSSE] or a combination of all [secSSE model, Leonel].

Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that is not
yet captured by the other diversification models. An example of this
is the multiple-birth model [Laudanno, Bilderbeek & Etienne, in preparation], 
that is the first model to allow speciation to co-occur.

An open question for each new diversification model is if the increased
complexity is worth the extra computational resources. There are many
ways to answer this question. One possible experiment 
is to use a phylogeny of the new diversification model,
simulate its alignment and see how well our
current inference models are at retrieving the phylogeny.
If this difference is small enough, current models apparently
suffice.

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek & Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [X and Bouckaert et al., ????]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to valorize new diversification models.

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given phylogeny of which full information is provided.
This involves multiple steps, described in detail below.

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given phylogenies to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic \cite{janzen2015approximate} by default,
but also an custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus & Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first 10%
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[language=R, floatplacement=H]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}
\giovanni{is "usethis" supposed to be "devtools"?}
\richel{"usethis" is the next "devtools", "devtools" will be deprecated in the future}
To start using \verb;pirouette;, load its functions in the global namespace first:

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, a clock
model of the alignment and a birth-death tree prior
to be the generative tree prior.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example.

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters.

>>>>

Phylogenies are commonly used to explore evolutionary hypotheses.
The mode of diversification is one such hypothesis.
Phylogenies constructed from DNA alignments are commonly fit to
diversification models, or the other way around. For example,
[Etienne $\&$ Rosindell 2012] fit four bird phylogenies to the protracted
diversification model.
\giovanni{ALTERNATIVE VERSION: The development of new powerful inference tools, such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], provided us the possibility to build phylogenetic trees from genetic material extracted from extant organisms. This has constituted a big step forward in the general understanding on how species evolved (this can be improved). Having access to this new kind of data has given rise to a high interest in understanding what are the main drivers and modes for diversification processes. }

There are plenty of diversification models: constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], time-dependent [?], diversity 
dependent [\cite{etienne2011diversity}], protracted 
birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman $\&$ Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman $\&$ Etienne, in preparation]
punctuated equilibrium aka early burst [?Harmon],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].
\giovanni{
@Richel: I am wondering if it might be worth to give a brief explanation on the philosophy of the aforementioned models, to explain why we want to simulate the processes. In case, this could be a stub for it:
Such models usually rely on the assumption that a particular diversification process can be explained mainly by specific mechanisms.
Such mechanisms are built into likelihood functions, which provide the likelihood of the parameters, given the data.
By maximizing the likelihood is possible to give the best inference of such parameters. It is therefore a standard practice, to validate such models, to test them versus simulated processes generated according to the hypothesised mechanisms.
}
Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that is not
yet captured by the other diversification models. An example of this
is the multiple-birth model [Laudanno, Bilderbeek, Haegeman $\&$ Etienne, in preparation], 
that is the first model to allow speciation to co-occur.

An open question for each new diversification model is whether the increased
complexity is worth the extra computational resources \giovanni{I think here it is not only a matter of computational resources. I believe there two more factors: 1) developing a likelihood model takes time, human resources and therefore fundings; 2) in principle you can elaborate any model to be as complex as possibly requested, however there is a diminishing return on this. In fact the purpose of modelling for complex systems such as biology is to hypothesize the main driving factors of a process and trying to prove that they are indeed able to explain the main features of the phenomenon you want to explain. In my opinion the job of a modeller is to try to disentangle the intrinsic complexity of a system highlighting the main reasons of its behaviour. In such a light you don't want your model to be unnecessary complex.}. There are many
ways to answer this question. One possible experiment 
is to use a phylogeny of the new diversification model,
simulate its alignment and see how well our
current inference models are at retrieving the phylogeny.
If this difference is small enough, current models apparently
suffice.

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek $\&$ Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to valorize new diversification models.

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given phylogeny of which full information is provided \giovanni{is this true? do we also provide information about extinct species or we just use reconstructed tree (which always lack a lot of information)?}.

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given phylogenies to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus $\&$ Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[language=R, floatplacement=H]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

To start using \verb;pirouette;, load its functions in the global namespace first:

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model \giovanni{do we have to explain what a site model is or what are the options?}, a clock
model of the alignment and a birth-death tree prior
to be the generative tree prior. \giovanni{would it be helpful to the reader to explicit that site and clock models are automatically taken from the generative model?}

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example \giovanni{it might be nice to add an explicit ref to the alignmentparams code}.

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters \giovanni{it might be nice to add an explicit ref to the lines of code we are referring to}.
